Open-source landscape for 3D CSEM modelling

### 2021-01-05 manuscript/Manuscript.tex
Add notes re feedback
Push in case there were changes.

% \documentclass[onecolumn,extra,referee,camera]{gji}
\documentclass[onecolumn,extra,camera]{gji}
\usepackage{timet}

% Additional packages
% \usepackage[UKenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
\usepackage{booktabs}
\usepackage{xspace}
\usepackage[pdftex, final, allcolors=blue, colorlinks=true]{hyperref}
\usepackage[pdftex]{graphicx}

@@

@@

% Own commands
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\emg}[2]{\texttt{emg#1#2}\xspace}
\newcommand{\empymod}{\texttt{empymod}\xspace}
\newcommand{\simpeg}{\texttt{SimPEG}\xspace}
\newcommand{\discretize}{\texttt{discretize}\xspace}
\newcommand{\custem}{\texttt{custEM}\xspace}
\newcommand{\petgem}{\texttt{PETGEM}\xspace}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}
\newcommand{\rmk}[1]{{\color{red}\bfseries #1}}

@@

@@

@@

In the 1990's, computers became sufficiently powerful that 3D modelling gained traction, and in 1995 the first International Symposium on Three-Dimensional Electromagnetics took place. This symposium resulted eventually in the book \emph{Three-Dimensional Electromagnetics} \citep{B.SEG.99.Oristaglio}, and another book by \cite{B.02.Wannamaker} with the exactly same title came out only three years later. Often cited publications from that time are \cite{RSC.94.Mackie} for 3D MT computation; \cite{RS.94.Druskin} for frequency- and time-domain modelling using a Yee grid and a global Krylov subspace approximation; and \cite{RS.96.Alumbaugh, GJI.97.Newman} for low-to-high frequency computation on massively parallel computers.

@@

Very well written overviews about the different approaches to 3D EM modelling (and inversion) are given by \cite{SG.05.Avdeev} and \cite{SG.10.Borner}. But there was also a tremendous publications output with regards to 3D EM modelling in the last 10-15 years, at least partly driven by the ever increasing computing power. One reason why there are so many publications about this topic results from the variety of techniques to solve the systems of linear equations (SLE). They can be distinguished between direct solvers \citep{GEO.09.Streich, GP.14.Chung, GEO.14.Jaysaval, GEO.15.Grayver, SEG.15.Oh, GJI.18.Wang}, indirect solvers \citep{GP.06.Mulder, GJI.15.Jaysaval}, or a combination of both, so-called hybrid solvers \citep{GEO.18.Liu}. The solvers often use preconditioners such as the multigrid method \citep{SIAM.02.Aruliah, GJI.16.Jaysaval}.

@@

The four codes under consideration are, in alphabetical order, \custem \citep{GEO.19.Rochlitz}, \emg3d \citep{JOSS.19.Werthmuller}, \petgem \citep{CAG.18.CastilloReyes, GJI.19.CastilloReyes}, and \simpeg \citep{CAG.15.Cockett, CAG.17.Heagy}. All four codes have their user-facing routines written in Python, and all of them make heavy use of \texttt{NumPy} \citep{NAT.20.Harris} and \texttt{SciPy} \citep{NM.20.Virtanen}. The four of them follow the open model of development, meaning that they come with both, an open-source license and an online-hosted version-control system with tracking possibilities (raising issues, filing pull requests). All developments comprise an extensive online documentation with many examples and have continuous integration to some degree. Newer package-management systems such as \texttt{conda}, \texttt{docker}, or \texttt{pip} simplify installation for all of these codes on any major operating system

@@

The Marlim oil field is a giant reservoir in a turbidite sandstone horizon in the north-eastern part of the Campos Basin, offshore Brazil, which was discovered in 1985. \cite{BJG.17.Carvalho} created from seismic data and well log data a realistic, three-dimensional resistivity model with vertical transverse isotropy (VTI), called MR3D, which they released under the open creative common license \emph{CC BY 4.0}. \cite{GEO.19.Correa} computed CSEM data for MR3D for six frequencies from 0.125\,Hz to 1.25\,Hz, and released them under the same CC license. To compute the data they used a state-of-the-art code from the industry \citep[][ \emph{SBLwiz} software from \emph{EMGS}]{GEO.07.Maao}. It is therefore, on one hand, an ideal case to validate our open-source codes against, as it is a complex, realistic model and the data were computed by an industry-proofed code. On the other hand it is impossible to reproduce it exactly, as it is a closed-source code and we cannot know exactly what was done internally. Additionally, that code is a time-domain code obtaining the frequency-domain responses through a transform, and it includes the air layer via a non-local boundary condition at the water-air interface \citep{GEO.10.Mittet}. We compute all the results in this study in the frequency domain, where we model the entire domain including the air layer by putting the boundaries of the computational domain far away.

The full MR3D model consists of $1022 \times 371 \times 1229$ cells, totalling to almost 466 million cells, where each cell has dimensions of 25 by 75 by 5\,m. The model was upscaled for the computation to 515 by 563 by 310 cells, totalling to almost 90 million cells, where each cell has dimensions of 100 by 100 by 20\,m. Both the full model and the upscaled computational model are released openly. The published data set consists of a regular grid of receivers of 20 in eastern direction by 25 in northern direction, 500 receivers in total, with 1\,km spacing located on the irregular seafloor. 45 source-towlines were located on the same grid above each receiver line, 50\,m above the seafloor, with shots every 100\,m. The computed responses for one of the receivers, \texttt{04Rx251a}, are shown in the original publication for the East-West inline-source \texttt{04Tx013a} and the East-West broadside-source \texttt{04Tx014a} (broadside offset of 1\,km). We reproduce the responses for this receiver and corresponding source-lines in our comparison. The $x$-$z$ cross-section of the horizontal resistivity model at the receiver position is shown in Figure~\ref{fig:model-marlim}, together with the receiver and sources positions. All layer have a VTI with $\lambda=\sqrt{2}$, except for the air layer, the seawater, and the salt layer, which are all isotropic.

@@

The actual differences in meshing is shown in the next two figures. Figure~\ref{fig:results-marlim_big} shows on the top row the rectilinear mesh of \emg3d, on the middle row the octree mesh of \simpeg, and on the bottom row the tetrahedral mesh of \custem and \petgem. The actual mesh and the broadside $|E_x|$ and $|E_y|$ fields are shown in the left, middle, and right columns, respectively. The extent of the plot includes the entire computational domain for the tetrahedral mesh (in x and z direction); the rectilinear and octree mesh extents are larger than what is shown (rectilinear: $x=327$ to 454\,km, $z=-57$ to 66\,km; octree: $x=288$ to 492\,km, $z=-123$ to 41\,km).
%
\begin{figure*}
  \centering
  \includegraphics[width=.9\linewidth]{figures/results-marlim_big}
  \caption{Meshes and broadside $E_x$ and $E_y$ fields in the left, middle, and right column, respectively, for the rectilinear, octree, and tetrahedral meshes in the top, middle, and bottom row, respectively. The shown dimensions shows the entire computational domain for the tetrahedra mesh, whereas the entire rectilinear and octree meshes are bigger.}
  \label{fig:results-marlim_big}
\end{figure*}
%
This figure is interesting for various points, as the differences between the three meshes are clearly visible. Particularly the advantage of octree and tetrahedra to create large cells outside of the region of interest to push away the boundary can be seen, whereas a rectilinear grid, even though stretched, has to compute many cells far away of the zone of interest. The different boundary condition also play a part, where the top row used PEC, and the other two rows PMC.

@@

%% 1. Results: good agreement %%
Overall, we observed an excellent match between the solutions. The amplitudes of the layered model have mostly a relative error of less than 1--2\,\%. Looking individually at the real or imaginary part results in much higher error values, but those are related to the zero crossings of the responses. The cross-validation of the results for the block model yields a similar picture, with normalised differences within a few percents. For the large, complex MR3D model it is a bit different. First it can be noted that the misfits are higher for higher frequencies. For the lowest frequency, 0.125\,Hz, the normalised difference is a few percent for inline and broadside $E_x$ components, and up to 10\,\% for the broadside $E_y$ component. For higher frequencies these numbers rise to the order of $\pm10\,$\%. The reason for the higher misfit for MR3D than for the layered or block model lies mostly in the discretization, boundary effects, and interpolation strategy. While one might argue that 5--10\,\% NRMSD is a lot, we argue that this is probably as good as it gets, given that the code with which the responses we compare to were computed is not open-source. This has been shown within the process of this study. Our first attempts were based on the originally published, fine MR3D model. This model is very detailed, but its extent is too small in x- and y-directions for CSEM computations. In order to represent the fine-scale model on a mesh suitable for computation, each of our codes did its own upscaling and extrapolation. However, we were unsuccessful in our attempt to reproduce the published results.  The reason is that there are many ways how you can extrapolate a model, and without knowing the details of how MR3D represents the model on the computational mesh, we could not re-produce this step. We reached out to the authors of MR3D, and they kindly agreed to also published the upscaled and extended computational model. It is only with this model that we were able to obtain comparable results.

@@

%% 3. Iterative vs direct solvers; runtime and memory %%
Whereas proper discretization is the main driver for accurate results, the used solver is the main discriminator in terms of runtime and memory. Both iterative and direct solvers are used in this study, however, the chosen models clearly favour iterative solvers. The biggest advantage of iterative solvers are the comparatively small memory requirements, even more so if the iterative solver is matrix free. Direct solvers exhibit their strongest advantage, besides stability and robustness, in terms of computation times if responses of multiple CSEM transmitters need to be computed in the same computational domain. As the system matrix factorisation requires 98--99\,\% of the solution time, computations for additional sources come at almost no cost for direct solvers, whereas it would come at the same cost as the first source for iterative solvers. Independently of the solver there is a trade-off between high accuracy and runtime. As an example, the octree mesh made for a very fast computation in the MR3D example thanks to the low number of dof, even though a direct solver was used. But the octree mesh has, due to its 2:1 aspect ratio, some limitations on the accuracy, why it was not used in the first two examples. Having discussed the discretization and the solver it is important to state that the reported runtime and memory is just one of the aspects, and we would like to emphasise that neither was at the core of this comparison, but the validation of the results. As such, no special efforts were undertaken to minimise either, as this is an entire different task.

>>>>

\documentclass[extra, camera,%
    % onecolumn,   % comment for two columns
    % referee,     % uncomment for double lines
    % final,       % uncomment for changes
]{gji}
\usepackage{timet}

% Additional packages
% \usepackage[UKenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
\usepackage{booktabs}
\usepackage{xspace}
\usepackage[final]{graphicx}
\usepackage[position=top]{subfig}
\usepackage{ifthen}
\usepackage[addedmarkup=bf]{changes}
% \usepackage[final]{changes}
\usepackage[pdftex, final, allcolors=blue, colorlinks=true]{hyperref}

@@

% Figures
% To print widths put this in text:
% \the\textwidth
% \the\columnwidth
%
% Widths:   textwidth  columnwidth  tablewidth
%       1-c   504pt        252pt       504pt
%       2-c   504pt        240pt       240pt
%       ref   468pt        234pt       468pt
%
\newlength{\fwidth}
\newlength{\cwidth}
\newlength{\twidth}
\ifthenelse{\boolean{GJ@referee}}{%
  \setlength{\cwidth}{234pt}%
  \setlength{\fwidth}{468pt}%
  \setlength{\twidth}{468pt}}{%
  \setlength{\fwidth}{504pt}%
  \ifthenelse{\boolean{@twocolumn}}{%
    \setlength{\cwidth}{240pt}%
    \setlength{\twidth}{240pt}}{%
    \setlength{\cwidth}{252pt}%
    \setlength{\twidth}{504pt}%
  }
}
\graphicspath{{./figures/}}

% Change default colour of «changes» package
\definecolor{alizarin}{rgb}{0.82, 0.1, 0.26}
\definecolor{mygreen}{rgb}{0, .5, 0}
\setaddedmarkup{\color{mygreen}{\uline{#1}}}
\setdeletedmarkup{\color{alizarin}{\uwave{#1}}}

@@

% Own commands
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\emg}[2]{\texttt{emg#1#2}\xspace}
\newcommand{\empymod}{\texttt{empymod}\xspace}
\newcommand{\simpeg}{\texttt{SimPEG}\xspace}
\newcommand{\discretize}{\texttt{discretize}\xspace}
\newcommand{\custem}{\texttt{custEM}\xspace}
\newcommand{\petgem}{\texttt{PETGEM}\xspace}

% Remove once done
% \newcommand{\itodo}[1]{\todo[color=blue!40!white, inline]{\sffamily #1}}
% \newcommand{\mtodo}[1]{\todo[inline]{\sffamily #1}}
% \newcommand{\etodo}[1]{\todo[color=blue!40!white]{#1}}
\newcommand{\dul}[1]{\underline{\underline{#1}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}

@@

% % == Notes from Wim; to incorporate ==
%
% All implemented
%
% % == Notes form Paulo ==
%
% Very good to know that the Ex responses are well represented. For most of the inversions done in the industry, we use these fields only.
%
% Ez: the mismatch occurs at the longer offsets and below the noise threshold 10-15. EMGS argues now that with the new source, the threshold is pushed to 10-16. But in practice, if the curve follows a smooth decay, we use this data for inversion, and we get meaningful geological results compared to seismic images.
%
% Ey: the broadside component is quite close,  with small differences at long offsets. For the inline component, I've noted that your responses are up to two orders of magnitude higher than MR3D. The four programs you tested give the same amplitudes? It may be a problem with the EMGS software, as EY is always neglected we may have ignored that issue. Simultaneously, it looks that your inline responses are at the same level as the broadside ones.
%
% Bathymetry is also a big issue. It must be well represented to avoid problems.
%
% % == Notes from Evert ==
%
% [T]he fact that the solutions in the four codes are roughly correct with an expected error between 1\% and 10\% makes it totally understandable that the field that is one to two orders of magnitude smaller has no accuracy. You could test this by driving the overall error down by a factor 100. Not sure that is feasible with all codes but if you can do it with one it can be enough when all codes agree sufficiently on the strong field components. Another option can be to increase the cross line component in the error computation.

@@

\begin{itemize}
  \item \deleted{old}: \texttt{\textbackslash deleted\{old\}}
  \item \added{new}: \texttt{\textbackslash added\{new\}}
  \item \replaced{new}{old}: \texttt{\textbackslash replaced\{new\}\{old\}}
\end{itemize}

@@

In the \replaced{1990s}{1990's}, computers became sufficiently powerful that 3D modelling gained traction, and in 1995 the first International Symposium on Three-Dimensional Electromagnetics took place. This symposium resulted eventually in the book \emph{Three-Dimensional Electromagnetics} \citep{B.SEG.99.Oristaglio}, and another book by \cite{B.02.Wannamaker} with the exactly same title came out only three years later. Often cited publications from that time are \cite{RSC.94.Mackie} for 3D MT computation; \cite{RS.94.Druskin} for frequency- and time-domain modelling using a Yee grid and a global Krylov subspace approximation; and \cite{RS.96.Alumbaugh, GJI.97.Newman} for low-to-high frequency computation on massively parallel computers.

@@

Very well written overviews about the different approaches to 3D EM modelling (and inversion) are given by \cite{SG.05.Avdeev} and \cite{SG.10.Borner}. But there was also a tremendous publications output with regards to 3D EM modelling in the last 10-15 years, at least partly driven by the ever increasing computing power. One reason why there are so many publications about this topic results from the variety of techniques to solve the systems of linear equations (SLE). They can be distinguished between direct solvers \citep{GEO.09.Streich, GP.14.Chung, GEO.14.Jaysaval, GEO.15.Grayver, SEG.15.Oh, GJI.18.Wang}, \replaced{iterative}{indirect} solvers \citep{GP.06.Mulder, GJI.15.Jaysaval}, or a combination of both, so-called hybrid solvers \citep{GEO.18.Liu}. The solvers often use preconditioners such as the multigrid method \citep{SIAM.02.Aruliah, GJI.16.Jaysaval}.

@@

The four codes under consideration are, in alphabetical order, \custem \citep{GEO.19.Rochlitz}, \emg3d \citep{JOSS.19.Werthmuller}, \petgem \citep{CAG.18.CastilloReyes, GJI.19.CastilloReyes}, and \simpeg \citep{CAG.15.Cockett, CAG.17.Heagy}. All four codes have their user-facing routines written in Python, and all of them make heavy use of \texttt{NumPy} \citep{NAT.20.Harris} and \texttt{SciPy} \citep{NM.20.Virtanen}. The four of them follow the open model of development, meaning that they come with both\deleted{,} an open-source license and an online-hosted version-control system with tracking possibilities (raising issues, filing pull requests). All developments comprise an extensive online documentation with many examples and have continuous integration to some degree. Newer package-management systems such as \texttt{conda}, \texttt{docker}, or \texttt{pip} simplify installation for all of these codes on any major operating system

@@

The Marlim oil field is a giant reservoir in a turbidite sandstone horizon in the north-eastern part of the Campos Basin, offshore Brazil, which was discovered in 1985. \cite{BJG.17.Carvalho} created from seismic data and well log data a realistic, three-dimensional resistivity model with vertical transverse isotropy (VTI), called MR3D, which they released under the open creative common license \emph{CC BY 4.0}. \cite{GEO.19.Correa} computed CSEM data for MR3D for six frequencies from 0.125\,Hz to 1.25\,Hz, and released them under the same CC license. To compute the data they used a state-of-the-art code from the industry \citep[][ \emph{SBLwiz} software from \emph{EMGS}]{GEO.07.Maao}. It is therefore, on \added{the} one hand, an ideal case to validate our open-source codes against, as it is a complex, realistic model and the data were computed by an industry-proofed code. On the other hand it is impossible to reproduce it exactly, as it is a closed-source code and we cannot know exactly what was done internally. Additionally, that code is a time-domain code obtaining the frequency-domain responses through a transform, and it includes the air layer via a non-local boundary condition at the water-air interface \citep{GEO.10.Mittet}. We compute all the results in this study in the frequency domain, where we model the entire domain including the air layer by putting the boundaries of the computational domain far away.

The full MR3D model consists of $1022 \times 371 \times 1229$ cells, totalling to almost 466 million cells, where each cell has dimensions of 25 by 75 by 5\,m. The model was upscaled for the computation to 515 by 563 by 310 cells, totalling to almost 90 million cells, where each cell has dimensions of 100 by 100 by 20\,m. Both the full model and the upscaled computational model are released \replaced{as open source}{openly}. The published data set consists of a regular grid of receivers of 20 in eastern direction by 25 in northern direction, 500 receivers in total, with 1\,km spacing located on the irregular seafloor. 45 source-towlines were located on the same grid above each receiver line, 50\,m above the seafloor, with shots every 100\,m. The computed responses for one of the receivers, \texttt{04Rx251a}, are shown in the original publication for the East-West inline-source \texttt{04Tx013a} and the East-West broadside-source \texttt{04Tx014a} (broadside offset of 1\,km). We reproduce the responses for this receiver and corresponding source-lines in our comparison. The $x$-$z$ cross-section of the horizontal resistivity model at the receiver position is shown in Figure~\ref{fig:model-marlim}, together with the receiver and sources positions. All layer have a VTI with $\lambda=\sqrt{2}$, except for the air layer, the seawater, and the salt layer, which are all isotropic.

@@

The actual differences in meshing is shown in the next two figures. Figure~\ref{fig:results-marlim_big} shows on the top row the rectilinear mesh of \emg3d, on the middle row the octree mesh of \simpeg, and on the bottom row the tetrahedral mesh of \custem and \petgem. The actual mesh and the broadside $|E_x|$ and $|E_y|$ fields are shown in the left, middle, and right columns, respectively. The extent of the plot includes the entire computational domain for the tetrahedral mesh (in x and z direction); the rectilinear and octree mesh extents are larger than what is shown (rectilinear: $x=327$ to 454\,km, $z=-57$ to 66\,km; octree: $x=288$ to 492\,km, $z=-123$ to 41\,km).
%
\begin{figure*}
  \centering
  \includegraphics[width=.9\linewidth]{figures/results-marlim_big}
  \caption{Meshes and broadside $E_x$ and $E_y$ fields in the left, middle, and right column, respectively, for the rectilinear, octree, and tetrahedral meshes in the top, middle, and bottom row, respectively. The shown dimensions \replaced{display}{shows} the entire computational domain for the tetrahedra mesh, whereas the entire rectilinear and octree meshes are bigger.}
  \label{fig:results-marlim_big}
\end{figure*}
%
This figure is interesting for various \replaced{reasons}{points}, as the differences between the three meshes are clearly visible. Particularly the advantage of octree and tetrahedra to create large cells outside of the region of interest to push away the boundary can be seen, whereas a rectilinear grid, even though stretched, has to compute many cells far away of the zone of interest. The different boundary condition also play a part, where the top row used PEC, and the other two rows PMC.

@@

%% 1. Results: good agreement %%
Overall, we observed an excellent match between the solutions. The amplitudes of the layered model have mostly a relative error of less than 1--2\,\%. Looking individually at the real or imaginary part results in much higher error values, but those are related to the zero crossings of the responses. The cross-validation of the results for the block model yields a similar picture, with normalised differences within a few percents. For the large, complex MR3D model it is a bit different. First it can be noted that the misfits are higher for higher frequencies. For the lowest frequency, 0.125\,Hz, the normalised difference is a few percent for inline and broadside $E_x$ components, and up to 10\,\% for the broadside $E_y$ component. For higher frequencies these numbers \replaced{increase}{rise} to the order of $\pm10\,$\%. The reason for the higher misfit for MR3D than for the layered or block model lies mostly in the discretization, boundary effects, and interpolation strategy. While one might argue that 5--10\,\% NRMSD is a lot, we argue that this is probably as good as it gets, given that the code with which the responses we compare to were computed is not open-source. This has been shown within the process of this study. Our first attempts were based on the originally published, fine MR3D model. This model is very detailed, but its extent is too small in \added{the} x- and y-directions for CSEM computations. In order to represent the fine-scale model on a mesh suitable for computation, each of our codes did its own upscaling and extrapolation. However, we were unsuccessful in our attempt to reproduce the published results.  The reason is that there are many ways how you can extrapolate a model, and without knowing the details of how MR3D represents the model on the computational mesh, we could not re-produce this step. We reached out to the authors of MR3D, and they kindly agreed to also published the upscaled and extended computational model. It is only with this model that we were able to obtain comparable results.

@@

%% 3. Iterative vs direct solvers; runtime and memory %%
Whereas proper discretization is the main driver for accurate results, the used solver is the main discriminator in terms of runtime and memory. Both iterative and direct solvers are used in this study, however, the chosen models clearly favour iterative solvers. The biggest advantage of iterative solvers are the comparatively small memory requirements, even more so if the iterative solver is matrix free. Direct solvers exhibit their strongest advantage, besides stability and robustness, in terms of computation times if responses of multiple CSEM transmitters need to be computed in the same computational domain. As the system matrix factorisation requires 98--99\,\% of the solution time, computations for additional sources come at almost no cost for direct solvers, whereas it would come at the same cost as the first source for iterative solvers. Independently of the solver there is a trade-off between high accuracy and runtime. \replaced{For}{As an} example, the octree mesh made for a very fast computation in the MR3D example thanks to the low number of dof, even though a direct solver was used. But the octree mesh has, due to its 2:1 aspect ratio, some limitations on the accuracy, why it was not used in the first two examples. Having discussed the discretization and the solver it is important to state that the reported runtime and memory is just one of the aspects, and we would like to emphasise that neither was at the core of this comparison, but the validation of the results. As such, no special efforts were undertaken to minimise either, as this is an entire different task.
